---
title: "AMS394_note2"
author: "Weihao Wang"
date: "2021/5/2"
output: html_document
---

**1. Probability and distributions**

1.1. Random sampling

The basic notion of a random sample is to deal from a well-shuffled pack of cards or picking numbered balls from a well-stirred urn.
In R, you can simulate these situations with the sample function. If you want to pick five numbers at random from the set 1:40, then you can write
```{r}
sample(1:40,5)
sample(c("H","T"), 10, replace=T)
sample(c("succ", "fail"), 10, replace=T, prob=c(0.9, 0.1))
```

1.2. Probability calculations and combinatorics

In R, the choose function can be used to calculate the number of ways to choose 5 numbers out of 40.
```{r}
choose(40,5)
prod(40:36)/prod(1:5)
factorial(40)/factorial(5)/factorial(35)
```

1.3. Discrete and continuous distributions

Discrete: Binomial distribution (binom), geometric distribution (geom), Poisson distribution (pois), ...
Continuous: Normal (norm), Beta (beta), Gamma (gamma), log-normal (lnorm), Cauchy (cauchy), Chi-squared (chisq), Exponential (exp), F (f), Logistic (logis), Negative binomial (nbinom), Student's t (t), Uniform (unif), ...

Prefix the name given here by 'd' for the density, 'p' for the CDF, 'q' for the quantile function and 'r' for simulation (random deviates). The first argument is x for dxxx, q for pxxx, p for qxxx and n for rxxx. We next discuss and give some examples on these functions.

**Distributions in R**
beta: beta - rbeta dbeta pbeta qbeta
binomial: binom
cauchy: cauchy
Chi-square: chisq
exponential: exp
F: f
gamma: gamma
geometric: geom
log-normal: lnorm
logistic: logis
Negative binomial: nbinom
Normal: norm
Poisson: pois
Student's t: t
uniform: unif
Weibull: weibull

1) Densities
```{r}
x <- seq(-4,4,0.1)
plot(x,dnorm(x),type="l")
curve(dnorm(x), from=-4, to=4)

x <- seq(0,1,0.01) 
plot(x,dbeta(x,4,5),type="l") 
curve(dbeta(x,4,5), from=0, to=1)

x <- seq(0,100,1) 
plot(x,dchisq(x,40),type="l") 
curve(dchisq(x,40), from=0, to=100)

x <- seq(0,100,1) 
plot(x,dgamma(x,30),type="l") 
curve(dgamma(x,30), from=0, to=100)
```

For discrete distributions, where variables can take on only distinct values, it is preferable to draw a pin diagram, here for the binomial distribution with n = 50 and p = 0.33:
```{r}
x <- 0:50
plot(x,dbinom(x,size=50,prob=.33),type="h") 
#curve(dbinom(x,size=50,prob=.33), from=0, to=50)
x <- 0:50 
plot(x,dpois(x,10),type="h")
```

2) Cumulative distribution functions
d-density
p-cdf

```{r}
pnorm(160,mean=132,sd=13)
pbinom(16,size=20,prob=.5)
pbeta(0.8,4,5)
pchisq(100,40)
pchisq(10,40)
pgamma(100,30)
pgamma(10,30)
ppois(2,10)

x <- seq(-4,4,0.1) 
plot(x,pnorm(x),type="l") 
curve(pnorm(x), from=-4, to=4)

x <- 0:50
plot(x,pbinom(x,size=50,prob=.33)) 
#curve(pbinom(x,size=50,prob=.33), from=0, to=50)
```

3) Quantiles

If we have n normally distributed observations with the same mean μ and standard deviation sigma, then it is known that the average 'xbar' is normally distributed around μ with standard deviation sigma/sqrt(n). A 95% confidence interval for μ can be obtained as
xbar + sigma/sqrt(n) × N0.025 <= μ<= xbar + s/pn × N0.975 
where N0.025 is the 2.5% quantile in the normal distribution.
```{r}
qnorm(0.5) 
qnorm(0.5,1,2) 
qnorm(0.025)

xbar=83
sigma=12
n<-5
sem<-sigma/sqrt(n)
sem 

xbar+sem*qnorm(0.025) 
xbar+sem*qnorm(0.975) 
xbar-sem*qnorm(0.025)
qbinom(0.5,size=20,prob=.5) 

qbeta(0.5,4,5) 
qbeta(1,4,5)
qchisq(0.5,40)
qchisq(1,40)
qgamma(0.5,30)
qgamma(1,30)
qpois(0.5,10)
qpois(1,10)
```

4) Random numbers

Computer generates sequences of “pseudo-random” numbers, which for practical purposes behave as if they were drawn randomly
```{r}
rnorm(10,mean=7,sd=5)
rbinom(10,size=20,prob=.5) 
rbeta(10,4,5)
rchisq(30,40)
rgamma(50,30)
rpois(20,10)
```

**2. Descriptive statistics and graphics**

2.1. Summary statistics for a single group

It is easy to calculate simple summary statistics with R. Here is how to calculate the mean, standard deviation, variance, and median.
```{r}
x <- rnorm(50)
mean(x)
sd(x) # standard deviation
var(x) # variance
sd(x)^2
median(x)
quantile(x)
pvec <- seq(0,1,0.1)
quantile(x,pvec)
quantile(x,c(0.1,0.4))

library(ISwR)
data(juul)
juul
dim(juul)
head(juul)
juul[1:10,]
attach(juul)
names(juul)
mean(igf1)
igf1
mean(igf1,na.rm=T)
sum(igf1)
sum(igf1,na.rm=T)
sum(is.na(igf1)) # number of missing values
sum(!is.na(igf1)) # number of nonmissing values
summary(igf1) # important
summary(juul)

str(juul) # important
juul<-transform(juul,sex=factor(sex, labels=c("f","m")),menarche=factor(menarche, labels=c("no","yes")),tanner=factor(tanner, labels=c("one","two","three","four","five")))

juul$sex = factor(juul$sex, labels=c("f","m"))
str(juul)
```

2.2 Graphic display of distributions --- Histograms, empirical distributions, Q-Q plot, Boxplot

```{r}
x<-rnorm(50) 
hist(x)

x<-c(1.5,2.5,3.5,4.5,5.5,6.5,8.5,9.5,12.5) 
y<-c(5,7,12,2,1,4,14,2,3)
z<-rep(x,y)
brk<-c(0,1,2,3,5,7,9,10,11,13) 
hist(z,breaks=brk)

### empirical distribution function
x<-rnorm(100)
n <- length(x)
plot(sort(x),(1:n)/n,type="s",ylim=c(0,1)) 
plot(sort(x),(1:n)/n,type="l",ylim=c(0,1)) 
plot(x,pnorm(x))
```

Empirical cumulative distribution function is defined as the fraction of data smaller than or equal to x. That is , if x is the kth smallest observation, the the proportion k/n of the data is smaller than or equal to x.

```{r}
#qq plot 
qqnorm(x)
```

```{r}
# Box plot
data(IgM)
par(mfrow=c(1,2))
boxplot(IgM)
boxplot(log(IgM))
par(mfrow=c(1,1))
```

mfrow graphical parameter should read as” multiframe, rowwise, 1*2 layout” individual plots are organized in 1 row and 2 columns.
```{r}
par(mfrow=c(1,2))
boxplot(log(IgM))
boxplot(IgM)
par(mfcol=c(2,2))
boxplot(log(IgM))
boxplot(IgM)
boxplot(sin(IgM))
boxplot(cos(IgM))
par(mfrow=c(2,2))
boxplot(log(IgM))
boxplot(IgM)
boxplot(sin(IgM))
boxplot(cos(IgM))
par(mfrow = c(1,1))
```

2.3. Summary statistics by groups

```{r}
data(red.cell.folate)
attach(red.cell.folate)
red.cell.folate
xbar=tapply(folate,ventilation,mean)
s=tapply(folate,ventilation,sd)
n=tapply(folate,ventilation,length)
cbind(mean=xbar,std.dev=s,n=n)
data(juul)
tapply(igf1,tanner,mean)
tapply(igf1,tanner,mean,na.rm=T) #na.rm = T --- deal with those missing values
```

na.rm=T as a parameter to mean to make it exclude the missing values.

4.4. Graphics for grouped data

Histograms
```{r}
data(energy)
attach(energy)
expend.lean<-expend[stature=="lean"]
expend.obese<-expend[stature=="obese"]
par(mfrow=c(2,1))
hist(expend.lean,breaks=10,xlim=c(5,13),ylim=c(0,4),col="white")
hist(expend.obese,breaks=10,xlim=c(5,13),ylim=c(0,4),col="grey")
par(mfrow=c(1,1))
```

Parallel boxplot
```{r}
boxplot(expend~stature)
boxplot(expend.lean , expend.obese)
```

Stripcharts
```{r}
opar<-par(mfrow=c(2,2),mex=0.8,mar=c(3,3,2,1)+.1)
stripchart(expend~stature)
stripchart(expend~stature,method="stack")
stripchart(expend~stature,method="jitter")
stripchart(expend~stature,method="stack",jitter=0.03)
par(opar)  #reestablished
```

2.5. Tables

Categorical data are usually described in the form of tables. A two-way table can be entered as a matrix object.
```{r}
caff.marital <- matrix(c(652,1537,598,242,36,46,38,21,218, 327,106,67), nrow=3,byrow=T)
colnames(caff.marital) <- c("0","1-150","151-300",">300")
rownames(caff.marital) <- c("Married","Prev.married","Single")
caff.marital
```

Furthermore, you can name the row and column names as follows. This is particularly useful if you are generating many tables with similar classification criteria.
```{r}
names(dimnames(caff.marital)) <- c("marital","consumption") # 1-row, 2-col
caff.marital
```

Like any matrix, a table can be transposed with the t function:
```{r}
t(caff.marital)
```


Exercise: Construct the following table which summarize the number of people smoking and nonsmoking in a class.
    smoking   nonsmoking
male   23       45
female 34       54

```{r}
data(juul)
attach(juul)
table(sex) # 1 variable - counting
table(sex,menarche)
table(menarche,tanner)
table(sex,menarche,tanner)
table(tanner,sex)
margin.table(table(tanner,sex),1)
margin.table(table(tanner,sex),2)
prop.table(table(tanner,sex),1)
prop.table(table(tanner,sex),2)
table(tanner,sex)/sum(table(tanner,sex))  #grand total of the table
```

2.6. Graphical display of tables

1) barplot
```{r}
barplot(prop.table(t(caff.marital)),legend.text=colnames(caff.marital),col=c("white","blue","green","black "))
```

2) dotcharts
```{r}
dotchart(t(caff.marital))
dotchart(caff.marital)
```

3) Pie charts
```{r}
opar<- par (mfrow=c(2,2), mex=0.8, mar=c(1,1,2,1))
slices<- c("white", "grey80", "grey50", "black") 
pie(caff.marital["Married",], main="Married", col=slices) 
pie(caff.marital["Prev.married",], main="Previously married", col=slices) 
pie(caff.marital["Single",], main="Single", col=slices)
par(opar)
```
